---
title: "CITE-seq sample assignment"
output:
  html_document:
    code_folding: show
    theme: cosmo
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
source("config/paths.R")
source("analysis/preprocessing/config.R")
preprocessing <- preprocessing_paths()
knitr::opts_knit$set(root.dir = preprocessing$root)
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 7
)
```

# 1. Set-Up

Assign demultiplexed sample IDs to integrated LOY cells using the barcode files for each pool.

## 1.1 Libraries

```{r setup_libs}
library(Seurat)
library(dplyr)
library(qs)
library(stringr)
```

## 1.2 Define Paths and Sample Information

```{r setup_variables}
# --- Input/Output Files ---
INTEGRATED_QS_PATH <- preprocessing$integrated
OUTPUT_QS_PATH <- preprocessing$sample_assigned

# --- Data Locations ---
INDIVIDUAL_COUNTS_DIR <- preprocessing$sample_barcodes

# Define which samples belong to which pool
POOL_MAP <- list(
  "Pool1" = c("S1", "S4"),
  "Pool2" = c("S5", "S6"),
  "Pool3" = c("S2", "S3")
)
```

# 2. Load Data

```{r load_integrated_object}
if (!file.exists(INTEGRATED_QS_PATH)) {
  stop(paste("Input file not found:", INTEGRATED_QS_PATH))
}

integrated_seurat_obj <- qread(INTEGRATED_QS_PATH)
cat("Successfully loaded the integrated Seurat object.\n")
```

# 3. Add New Metadata with Strict Mapping

Map each pool using only its constituent sample barcode files.

```{r add_metadata_strict}
cat("\n--- Adding 'sample_id' metadata using a strict, pool-specific mapping logic ---\n")

# Get the full metadata data frame
metadata <- integrated_seurat_obj@meta.data

# Create the new 'sample_id' column, initializing with NA
metadata$sample_id <- NA_character_

# Loop through each pool defined in our map
for (pool_name in names(POOL_MAP)) {

  samples_in_pool <- POOL_MAP[[pool_name]]
  cat(paste0("\nProcessing ", pool_name, " (contains samples: ", paste(samples_in_pool, collapse=", "), ")\n"))

  # --- Create a barcode map ONLY for the samples in this pool ---
  pool_barcode_map <- character()
  for (sample_name in samples_in_pool) {
    barcode_file_path <- file.path(INDIVIDUAL_COUNTS_DIR, sample_name, "sample_filtered_feature_bc_matrix", "barcodes.tsv.gz")
    if (!file.exists(barcode_file_path)) {
      stop(paste("Barcode file not found for sample:", sample_name))
    }
    barcodes <- read.table(barcode_file_path, header = FALSE)$V1
    sample_map <- setNames(rep(sample_name, length(barcodes)), barcodes)
    pool_barcode_map <- c(pool_barcode_map, sample_map)
  }

  # --- Apply this pool-specific map ---
  # Identify which cells in the full object belong to this pool
  is_in_current_pool <- metadata$orig.ident.x %in% pool_name

  if(sum(is_in_current_pool) == 0) {
      cat(paste("Warning: No cells found for", pool_name, "in the Seurat object. Skipping.\n"))
      next
  }

  # Get the cell names for only the cells in this pool
  pool_cell_names <- rownames(metadata)[is_in_current_pool]

  # Extract the core barcode sequence
  core_barcodes <- str_remove(pool_cell_names, "^[a-zA-Z0-9]+_")

  # Use the pool-specific map to look up the sample ID
  sample_ids_for_pool_cells <- unname(pool_barcode_map[core_barcodes])

  # Fill in the 'sample_id' column for only the cells in this pool
  metadata$sample_id[is_in_current_pool] <- sample_ids_for_pool_cells

  cat(paste("Finished mapping cells for", pool_name, "\n"))
}

# Add the updated metadata back to the Seurat object
integrated_seurat_obj$sample_id <- metadata$sample_id

cat("\nSuccessfully added the 'sample_id' metadata column.\n")
```

# 4. Verification

Summarize sample assignments within each pool.

```{r verify_metadata}
cat("\n--- Verifying the new metadata ---\n\n")

cat("Total cell counts per original sample (S1-S6 should have counts, others NA):\n")
print(table(integrated_seurat_obj$sample_id, useNA = "ifany"))

cat("\nCell counts per sample within each pool:\n")
# Filter metadata for only LOY pools before creating the table
loy_meta <- integrated_seurat_obj@meta.data %>% filter(!is.na(sample_id))
print(table(Pool = loy_meta$orig.ident.x, Sample = loy_meta$sample_id))
```

# 5. Save Updated Object

Save the object with the assigned sample IDs.

```{r save_updated_object}
# Save the final integrated object with the new metadata
qs::qsave(integrated_seurat_obj, OUTPUT_QS_PATH)
cat(paste("\nSuccessfully saved the updated Seurat object to:", OUTPUT_QS_PATH, "\n"))
```

```{r sessionInfo}
sessionInfo()

```
