---
title: "CITE-seq cell annotation"
output:
  html_document:
    code_folding: show
    theme: cosmo
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
source("config/paths.R")
source("analysis/preprocessing/config.R")
preprocessing <- preprocessing_paths()
knitr::opts_knit$set(root.dir = preprocessing$root)
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 7
)
```

# 1. Set-Up

Inspect RNA and ADT markers, apply the cluster annotation map, and summarize cell-type composition.

## 1.1 Libraries

```{r setup_libs}
library(Seurat)
library(ggplot2)
library(patchwork)
library(dplyr)
library(ggthemes)
library(qs)

```

## 1.2 Load Data and Parameters

```{r setup_variables}
# --- Input File ---
# Path to the integrated Seurat object saved from the previous step
INPUT_QS_PATH <- preprocessing$sample_assigned

# --- Analysis Parameters ---
# List of antibodies (must match previous scripts)
ANTIBODY_LIST <- c("CD123", "CD34", "CD33", "HLA-DR", "TCRab", "CD4", "CD19", "CD8", "CD3")

# List of canonical gene markers to visualize
GENE_MARKER_LIST <- c(
  # Progenitor / AML Blast markers
  "CD34", "SOX4", "GATA2",
  # Myeloid markers
  "MPO", "ELANE", "LYZ", "CD33", "CEBPE",
  # T-Cell markers
  "CD3E", "CD4", "CD8A",
  # B-Cell markers
  "MS4A1", "CD19", "JCHAIN",
  # NK Cell markers
  "NKG7", "GNLY"
)

# Set a consistent theme for plots
theme_set(theme_clean())

```

```{r load_data}
if (!file.exists(INPUT_QS_PATH)) {
  stop(paste("Input file not found:", INPUT_QS_PATH))
}

# Load the final integrated Seurat object
integrated_seurat_obj <- qread(INPUT_QS_PATH)

cat("Successfully loaded the integrated Seurat object.\n")
print(integrated_seurat_obj)
```

# 2. UMAP Visualizations

## 2.1 UMAP by Sample and Cluster

Visualize the integrated embedding by sample and cluster.

```{r umap_by_sample_cluster, fig.width=14, fig.height=6}
# Set default assay to RNA for plotting gene expression later
DefaultAssay(integrated_seurat_obj) <- "RNA"

p_by_sample <- DimPlot(integrated_seurat_obj, reduction = "umap", group.by = "orig.ident", pt.size = 0.2) +
               labs(title = "Integrated UMAP by Sample Pool")

p_by_cluster <- DimPlot(integrated_seurat_obj, reduction = "umap", label = TRUE, repel = TRUE, pt.size = 0.2) +
                labs(title = "Integrated UMAP by Cluster") + NoLegend()

p_by_sample + p_by_cluster
```

# 3. Marker Visualization

Inspect protein and RNA markers across the integrated embedding.

## 3.1 Antibody (ADT) Feature Plots

Inspect ADT expression across cell populations.

```{r adt_feature_plots, fig.width=12, fig.height=10}
# Set default assay to ADT for plotting protein expression
DefaultAssay(integrated_seurat_obj) <- "ADT"

# Use ADT feature names stored in the object.
adt_features_in_obj <- rownames(integrated_seurat_obj[["ADT"]])

FeaturePlot(
  integrated_seurat_obj,
  features = adt_features_in_obj,
  reduction = "umap",
  min.cutoff = "q05",
  max.cutoff = "q95",
  pt.size = 0.1,
  cols = c("lightgrey", "darkred"),
  ncol = 3
)
```

## 3.2 Canonical Gene Marker Feature Plots

Inspect canonical RNA markers for cell-type annotation.

```{r rna_feature_plots, fig.width=12, fig.height=15}
DefaultAssay(integrated_seurat_obj) <- "RNA"

# Filter for genes that are actually in the object to avoid errors
gene_markers_in_obj <- intersect(GENE_MARKER_LIST, rownames(integrated_seurat_obj))

FeaturePlot(
  integrated_seurat_obj,
  features = gene_markers_in_obj,
  reduction = "umap",
  min.cutoff = "q05",
  max.cutoff = "q95",
  pt.size = 0.1,
  cols = c("lightgrey", "darkblue"),
  ncol = 3
)
```

# 4. Cluster Marker Identification

Identify positive RNA markers for each cluster.

## 4.1 Find All Cluster Markers (RNA)

```{r find_markers}
# Set Idents to the final clusters
Idents(integrated_seurat_obj) <- "seurat_clusters"
DefaultAssay(integrated_seurat_obj) <- "RNA"
integrated_seurat_obj <- JoinLayers(integrated_seurat_obj)
# Find markers for every cluster compared to all remaining cells, report only the positive ones
cluster_markers <- FindAllMarkers(integrated_seurat_obj, only.pos = TRUE, min.pct = 0.25, logfc.threshold = 0.25, verbose = FALSE)

# Get the top 10 markers for display
top10_markers <- cluster_markers %>%
  group_by(cluster) %>%
  top_n(n = 10, wt = avg_log2FC)
```

## 4.2 Visualize Top Markers with a Dot Plot

Display the top cluster markers.

```{r dotplot_markers, fig.height=8, fig.width=12}
DotPlot(integrated_seurat_obj, features = unique(top10_markers$gene)) +
  labs(title = "Top 10 Marker Genes per Cluster") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

# 5. Cell-type annotation

```{r annotate_clusters}
# Use clustering labels as the starting identities
Idents(integrated_seurat_obj) <- integrated_seurat_obj$seurat_clusters

# Build a full mapping: default = "Leukemia", with explicit overrides
clust_levels <- levels(integrated_seurat_obj$seurat_clusters)
map_simple <- setNames(rep("Leukemia", length(clust_levels)), clust_levels)

# Cluster-specific cell-type assignments
map_simple[c("3")]                  <- "CD4 T"
map_simple[c("10")]                 <- "CD8 T"
map_simple[c("13")]                 <- "B cell"
map_simple[c("6","9","12","19","20","21","23")] <- "Unidentified"

# Apply mapping
integrated_seurat_obj <- RenameIdents(integrated_seurat_obj, map_simple)

# Save to metadata (consistent naming)
integrated_seurat_obj <- AddMetaData(
  integrated_seurat_obj,
  metadata = Idents(integrated_seurat_obj),
  col.name = "cell_type"
)

```

# 6. Cluster Proportions by Sample

Summarize cell-type composition by sample.

```{r cluster_proportions, fig.width=12, fig.height=6}
# Calculate proportions
cluster_props <- integrated_seurat_obj@meta.data %>%
  group_by(cell_type, orig.ident) %>%
  summarise(n = n(), .groups = 'drop') %>%
  group_by(orig.ident) %>%
  mutate(prop = n / sum(n))

# Plot
ggplot(cluster_props, aes(x = cell_type, y = prop, fill = orig.ident)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Proportion of Cells",
    x = "Cluster",
    y = "Proportion",
    fill = "Sample"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r sessionInfo}
sessionInfo()
```

```{r}
# cat1 = cell type (rows), cat2 = sample/group (columns)
pct_table <- function(cat1, cat2, within = c("sample", "celltype"), digits = 1) {
  within <- match.arg(within)
  tab <- table(cat1, cat2, dnn = c("cell_type", "group"))
  margin <- if (within == "sample") 2 else 1   # 2 = column-wise (per sample); 1 = row-wise (per cell type)
  round(prop.table(tab, margin = margin) * 100, digits)
}

# counts
count_table <- function(cat1, cat2) {
  table(cat1, cat2, dnn = c("cell_type", "group"))
}


# Plot
DimPlot(integrated_seurat_obj, reduction = "umap", label = TRUE, pt.size = 0.7)

# ---------------------------
# Counts and percentages by sample
# ---------------------------
print_total_pct <- function(celltype, by, digits = 1){
  tb  <- table(celltype, by)
  pct <- prop.table(tb, margin = 2) * 100
  cbind(Count = tb, Percent = round(pct, digits))
}

by_field <- if ("sample" %in% colnames(integrated_seurat_obj@meta.data)) {
  integrated_seurat_obj$sample
} else {
  integrated_seurat_obj$orig.ident
}

knitr::kable(table(integrated_seurat_obj$cell_type, by_field),
             caption = "Cell-type counts by sample")

knitr::kable(print_total_pct(integrated_seurat_obj$cell_type, by_field),
             caption = "Cell-type percentages by sample")

tab_counts <- count_table(integrated_seurat_obj$cell_type, by_field)
tab_pct    <- pct_table(integrated_seurat_obj$cell_type, by_field, within = "sample", digits = 1)

knitr::kable(tab_counts, caption = "Counts: cells per cell type by sample")
knitr::kable(tab_pct,    caption = "Percent of each sample that is each cell type")

```

```{r}

library(dplyr)
library(tidyr)
library(ggplot2)
library(forcats)
library(scales)
library(knitr)

# Work off metadata
meta <- integrated_seurat_obj@meta.data

# Counts per cell type x sample
cluster_counts <- meta %>%
  count(cell_type, orig.ident, name = "n")

# Totals per cell type (for ordering)
totals_ct <- cluster_counts %>%
  group_by(cell_type) %>%
  summarise(total = sum(n), .groups = "drop")

# Sample fractions within each cell type
cluster_props_ct <- cluster_counts %>%
  group_by(cell_type) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  left_join(totals_ct, by = "cell_type") %>%
  mutate(cell_type = fct_reorder(cell_type, total, .desc = TRUE))

# --- Plot 1: Percent of each cell type that comes from each sample ---
p_pct <- ggplot(cluster_props_ct, aes(x = cell_type, y = prop, fill = orig.ident)) +
  geom_col() +  # already normalized within cell type
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Sample composition",
    x = "Cell type",
    y = "Percent of cell type",
    fill = "Sample"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# --- Plot 2: Absolute counts per cell type stacked by sample ---
p_abs <- ggplot(cluster_props_ct, aes(x = cell_type, y = n, fill = orig.ident)) +
  geom_col(position = "stack") +
  labs(
    title = "Cell counts by type and sample",
    x = "Cell type",
    y = "Cell count",
    fill = "Sample"
  ) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Print the plots
print(p_pct)
print(p_abs)

# --- Tables (optional) ---
# Wide counts
count_table_wide <- cluster_counts %>%
  pivot_wider(names_from = orig.ident, values_from = n, values_fill = 0) %>%
  arrange(desc(rowSums(across(where(is.numeric)))))

# Wide percentages (within cell type)
pct_table_wide <- cluster_props_ct %>%
  mutate(Percent = round(prop * 100, 1)) %>%
  select(cell_type, orig.ident, Percent) %>%
  pivot_wider(names_from = orig.ident, values_from = Percent, values_fill = 0) %>%
  arrange(match(cell_type, levels(cluster_props_ct$cell_type)))

kable(count_table_wide, caption = "Counts per cell type by sample")
kable(pct_table_wide,   caption = "Percent within each cell type (by sample)")

qs::qsave(integrated_seurat_obj, preprocessing$annotated)
```
