---
title: "Shared genes and Y-chromosome gene retention"
output:
  html_document:
    code_folding: show
    theme: cosmo
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
source("config/paths.R")
source("analysis/preprocessing/config.R")
preprocessing <- preprocessing_paths()
knitr::opts_knit$set(root.dir = preprocessing$root)
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 10, fig.height = 7
)
```

# 1. Set-Up

Retain genes expressed in both cohorts, together with Y-chromosome genes retrieved from Ensembl.

## 1.1 Libraries

```{r setup_libs}
library(Seurat)
library(dplyr)
library(qs)
library(biomaRt)
```

## 1.2 Define Paths and Parameters

```{r setup_variables}
# --- Input/Output Files ---
INPUT_QS_PATH <- preprocessing$annotated
OUTPUT_QS_PATH <- preprocessing$shared_genes

# --- Cohort Definition ---
NEW_COHORT_IDS <- c("Pool1", "Pool2", "Pool3")
```

# 2. Load Integrated Seurat Object

```{r load_data}
if (!file.exists(INPUT_QS_PATH)) {
  stop(paste("Input file not found:", INPUT_QS_PATH))
}

integrated_seurat_obj <- qread(INPUT_QS_PATH)

cat("Successfully loaded the full integrated Seurat object.\n")
cat("Original object dimensions:\n")
print(integrated_seurat_obj)
```

# 3. Fetch All Y-Chromosome Genes

Retrieve Y-chromosome gene symbols with `biomaRt`.

```{r fetch_y_genes}
cat("\n--- Fetching all Y-chromosome genes from Ensembl... ---\n")
ensembl <- useMart("ensembl", dataset = "hsapiens_gene_ensembl")
y_genes_df <- getBM(
  attributes = c("hgnc_symbol"),
  filters = "chromosome_name",
  values = "Y",
  mart = ensembl
)

# Extract the gene symbols into a character vector, removing any blank entries
ALL_Y_GENES_FROM_ENSEMBL <- y_genes_df$hgnc_symbol
ALL_Y_GENES_FROM_ENSEMBL <- unique(ALL_Y_GENES_FROM_ENSEMBL[ALL_Y_GENES_FROM_ENSEMBL != ""])
cat(paste("Found", length(ALL_Y_GENES_FROM_ENSEMBL), "unique Y-chromosome gene symbols.\n"))
```

# 4. Identify Genes to Keep

Combine genes expressed in both cohorts with the Y-chromosome gene list.

```{r find_genes_to_keep}
cat("\n--- Identifying final gene set for filtering ---\n")

# Get the RNA counts matrix
rna_counts <- GetAssayData(integrated_seurat_obj, assay = "RNA", layer = "counts")

# --- 4a. Identify cells in each cohort ---
all_cell_barcodes <- colnames(integrated_seurat_obj)
new_cohort_cells <- all_cell_barcodes[integrated_seurat_obj$orig.ident.x %in% NEW_COHORT_IDS]
old_cohort_cells <- all_cell_barcodes[!(integrated_seurat_obj$orig.ident.x %in% NEW_COHORT_IDS)]

# --- 4b. Find the intersection of expressed genes ---
genes_in_new_cohort <- rownames(rna_counts)[rowSums(rna_counts[, new_cohort_cells, drop = FALSE]) > 0]
genes_in_old_cohort <- rownames(rna_counts)[rowSums(rna_counts[, old_cohort_cells, drop = FALSE]) > 0]
common_genes <- intersect(genes_in_new_cohort, genes_in_old_cohort)
cat(paste("Found", length(common_genes), "genes common to both cohorts.\n"))

# --- 4c. Combine common genes with ALL Y-chromosome genes ---
# This ensures that even if a Y-gene was not "common", it will be kept.
final_genes_to_keep <- unique(c(common_genes, ALL_Y_GENES_FROM_ENSEMBL))
cat(paste("Final gene list (common genes + all Y genes) contains", length(final_genes_to_keep), "genes.\n"))
```

# 5. Create New Seurat Object with Filtered Genes

```{r create_new_object}
cat("\n--- Creating new Seurat object with the filtered gene list ---\n")

# Subset the RNA assay to keep only the desired genes.
# Keep the original object and write the filtered object separately.
filtered_seurat_obj <- subset(integrated_seurat_obj, features = final_genes_to_keep)

cat("Successfully created the new, gene-filtered Seurat object.\n")
```

# 6. Verification

```{r verify_object}
cat("\n--- Verifying object dimensions ---\n")
cat("\nOriginal object dimensions:\n")
print(integrated_seurat_obj)
cat("\nNew filtered object dimensions:\n")
print(filtered_seurat_obj)

# Check if a key Y-gene like PRKY is still in the new object
prky_check <- "PRKY" %in% rownames(filtered_seurat_obj[["RNA"]])
cat(paste0("\nVerification check: Is 'PRKY' present in the new object? ", prky_check, "\n"))
if(!prky_check) {
    warning("Warning: PRKY was not found in the final object. Check gene names and biomaRt output.")
}
```

# 7. Save Filtered Object

```{r save_object}
qs::qsave(filtered_seurat_obj, OUTPUT_QS_PATH)
cat(paste("\nSuccessfully saved the filtered Seurat object to:", OUTPUT_QS_PATH, "\n"))
```

```{r sessionInfo}
sessionInfo()

```
